Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-02-09 second-draft/2-paper-names.tex
revising paper

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. In this work we examine this question from a computational perspective, demonstrating a fully formalized framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
`To' is expressed by a \emph{preposition}, so named because it precedes the word expressing `friend'. %The adpositional phrase starts with the adposition.
In Japanese, the verb stands at the end; object and adpositional phrase precede it; `to' is expressed by a \textit{postposition}, so named because it follows the word denoting `friend'.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb, and that the objects of adpositions and verbs stand on the sam side of their heads.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
%It turns out that these are deep and stable regularities:
It turns out that this variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions:
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.
%However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
This correlation has an extremely strong empirical basis:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

@@

%These cross-linguistic patterns call for an explanation, and 
Explaining these patterns has been an important aim of linguistic research since Greenberg's seminal study.
%and the problem of explaining the word order correlations has inspired a long tradition of research.

%An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.

@@

We formalize efficiency of language as a tradeoff between minimizing the

\cite{zipf1949human} argued that language optimizes a tradeoff between speaker effort (Force of Unification) and listener effort (Force of Diversification).
According to Zipf, a language that minimizes speaker effort should reduce the number of different utterances to make production as easy as possible.
Conversely, to minimize listener effort, the language should provide different utterances for different meanings, so that the listeenr can unambiguously identify the meaning from the utterance.
Minimizing speaker and listener represent two opposing forces:
A pressure to reduce the complexity of the language makes the utterances in a language more homogenous.
A pressure to reduce ambiguity makes utterances more heterogeneous, so they can indicate different meanings.
The idea that language results from the tension between these two pressures has a long and fruitful history.
\cite{horn-toward-1984}, RSA, Color naming

This idea has been shown to account for phenomena such as pragmatic inference (CITE) and color naming (CITE).

TODO somewhere start calling this Efficiency

TODO explain why surprisal. some production/speaker work. (plan reuse)

@@

In this work, we show computationally, using corpus data from 51 languages, that efficiency optimization accounts for the Greenbergian word order correlations.
That is, we show that the word orders in natural languages have evolved to optimize efficiency, and that this optimization accounts for the prevalence of the word order correlations. %word orders in natural language

@@

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|llllllll}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &         \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object     & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
	adposition    &    NP       & 	\emph{to}            & \emph{a friend} \\

copula    &    NP        &     \\
	\emph{	is}        & \emph{a friend} \\
auxiliary    &    VP       &   \\
	has          & written \\
noun    &    genitive      &  \\
friend &  of John \\
noun    &    relative clause       &  \\
	books & that you read \\
complementizer    &    S        & \\
	that & Mary \\
verb    &    PP     &    \\
	went & to school \\
want    &    VP      &    \\
	wants   & to to leave \\
verb    &    subject    &  \\
	(there) entered & a tall man \\
verb    &    manner adverb     & \\
	ran & quickly\\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
For each correlation, we provide an example.
Given the statistical nature of the correlations, not every language satisfies every one of them.
Not all correlations are satisfied by every natu
}\label{table:corr-dryer}
\end{table*}

@@

We model the process where a speaker communicates an utterance to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In Shannon's model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of a set of syntactic and semantic relations.
Following a long tradition in formal and computational liguistics, we formalize these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
In Figure~\ref{fig:comm}, the message consists of a dependency structure with the words `saw', `Mary', `John', where Mary is the subject of the event denoted by `saw', and John is the object.

@@

In Information Theory, the usefulness of a communication channel depends on (1) how costly encoding and transmission is, and (2) how precisely messages can be recovered from codes.
In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance (TODO can we find an example for this?)

The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances provide about their underlying tree structures:
\begin{equation}
	R_{Pars} := I[L,T] = \sum_{t,l} p(t,l) \log \frac{p(t|l)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
This quantity formalizes Zipf's principle of listener effort and his Force of Diversification:
It can be maximized when trees can be decoded fully unambiguously from utterances.
This formalization is a standard formalization of Zipf's principle of listener economy (CITE RSA, Regier, ...).

The counteracting Force of Unification is formalized by the cost of encoding the dependency structure and producing the utterance.
TODO motivate surprisal here.

@@

Do languages optimize their word order for efficiency?
Does this optimization explain the observed language universals?

@@

The \key{predictability} of the language is the negative entropy of the sentences:
\begin{equation}
	R_{Pred} := H[L] = \sum_{l} p(l) \log p(l)
\end{equation}
where the sum runs over all possible sentences $l$ that belong to the language.
This quantity describes how homogeneous the language is, i.e., it is larger if the distribution over sentences is concentrated on a smaller number of frequent sentences. %how many distinct sentences there are.
%It is larger

The opposing pressure is \key{parsability}, which is the mutual information beteen sentences and dependency structures:
\begin{equation}
	R_{Pars} := I[L,T] = \sum_{t,l} p(t,l) \log \frac{p(t|l)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.

The \key{efficiency} of a language is a weighted combination
\begin{equation}
	R_{Eff} := R_{Pars} + \lambda R_{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification).

@@

%The two terms of Eq.~\ref{eq:efficiency} implement two opposing pressures. Parseability (the mutual information term) pushes to make strings more heterogeneous, so that they can indicate different tree structures unambiguously. Predictability (the entropy term) pushes to make strings more homogeneous, to reduce the complexity of using the language.
%The idea that natural language arises from a tension between these two forces goes back over a century \cite{gabelentz1901sprachwissenschaft}.
Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking unordered dependency trees $T$ as the underlying meanings to be conveyed. Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship).

\subsection{Optimizing Grammars for Efficiency}

We now ask:
(1) Do the grammars of natural languages evolve towards optimizing efficiency of communication?
(2) Does this process of optimization account for Greenberg's order correlation universals?

@@

@@

\subsection{Efficiency Explains DLM}
Finally we demonstrate the relationship between dependency length minimization and the maximization of efficiency. Figure~\ref{fig:deplength} shows average dependency length per sentence length for four typologically distinct languages, showing real languages, random baselines, and languages optimized for dependency length, parseability, predictability, and efficiency. We see that optimizing for efficiency lowers dependency length relative to random baselines, in keeping with the suggestion that dependency length minimization is a by-product of efficiency maximization \cite{futrell2017memory}. In 80\% of the languages, optimizing explicitly for dependency length produces dependencies that overshoot the depenndency length of the real language; in 3/4 of the lanuages shown, the real language is best matched by efficiency optimization.

@@

@@

@@

See the SI for details on the neural language models and parsers used, and on the optimization procedures.

>>>>

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.

In this work we examine this question from a computational perspective, demonstrating a fully formalized framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

We aim to explain the linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. These universals are called \key{implicational} universals, and take the form of correlations between certain word order patterns and others across languages. For example, whether a language puts verbs before or after objects strongly determines whether it places genitives before or after nouns, and whether it has prepositions or postpositions \cite{dryer1992greenbergian}.

Researchers working in the functional paradigm have proposed a range of criteria which languages should meet in order to enable efficient communication and processing. These criteria are based on theories of online processing difficulty (see \cite{jaeger2011language} for a review) and on information theoretic notions of efficiency and robustness in communication \cite{ferrericancho2001two,piantadosi2011word,gibson2013noisy}, and they have been formalized to varying degrees.

We formalize the notion of communicative efficiency using two terms: one having to do with minimizing the difficulty of linguistic communication, and one having to do with maximizing the information transferred in such communication. We quantify the difficulty of linguistic communication in terms of the word-by-word \key{predictability} of strings, which has been found to be a highly accurate and general predictor of human online comprehension difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}, and which can be justified in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}. We quantify the information transferred in language as \key{parseability}: the quantity of information contained in linguistic utterances about latent hierarchical tree structures. We also compare the results of optimizing for communicative efficiency to the results of optimizing for dependency length minimization, a previously proposed approximate metric of the amount of memory involved in language production and comprehension \cite{hudson1995measuring,gibson1998linguistic,gibson2000dependency,liu2017dependency,temperley2018minimizing}.

%To evaluate whether communicative efficiency can explain word order universals, we apply the following method \cite{gildea2007optimizing}. First, we develop word order grammars: simple, interpretable models of word order, implemented as functions that take in an unordered syntactic dependency tree and output a distribution over possible word orders for that tree. Second, we take dependency treebanks of natural languages and compute word order grammars which are optimized for an objective functions representing communicative efficiency.
%Third, we evaluate whether the optimized word order grammars reproduce the desired word order universals when applied to attested dependency trees from the treebanks across many languages. Our favorable results suggest that it may be possible to generally model human languages as efficient codes for communication among agents with generic human-like cognitive and communicative constraints. %The best way to describe a human language might be as a solution to an equation involving such constraints.

\section{Greenberg's Word Order Correlations}
Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
`To' is expressed by a \emph{preposition}, so named because it precedes the word expressing `friend'. %The adpositional phrase starts with the adposition.
In Japanese, the verb stands at the end; object and adpositional phrase precede it; `to' is expressed by a \textit{postposition}, so named because it follows the word denoting `friend'.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb, and that the objects of adpositions and verbs stand on the sam side of their heads.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
%It turns out that these are deep and stable regularities:
It turns out that this variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions:
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.
%However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
This correlation has an extremely strong empirical basis:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

TODO maybe being more general in the beginning, as in the first version. make sure the reader doesn't think 'it's just headedness, why bother'

@@

%These cross-linguistic patterns call for an explanation, and 
%and the problem of explaining the word order correlations has inspired a long tradition of research.
Supported by languages on all continents, these correlations are among the language universals with the strongest empirical support.
Importantly, their validity is also independent from specific assumptions about theories of grammar.

Consequently, explaining these patterns has been an important aim of linguistic research since Greenberg's seminal study.
%
%%supported by languages on all continents, and their validity is independent from 
%
%TODO say more about why these are extremely important, and that it is a puzzle why they are true
%
%%An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. 
%
A prominent line of research has argued that universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

@@

Zipf \cite{zipf1949human} argued that language optimizes a tradeoff between speaker effort (Force of Unification) and listener effort (Force of Diversification).
According to this theory, a language that minimizes speaker effort should reduce the number of different utterances to make production as easy as possible.
Conversely, to minimize listener effort, the language should provide different utterances for different meanings, so that the listeenr can unambiguously identify the meaning from the utterance.
Minimizing speaker and listener effort represent two opposing forces:
A pressure to reduce the complexity of the language makes the utterances in a language more homogenous.
A pressure to reduce ambiguity makes utterances more heterogeneous, so they can indicate different meanings.
%Another way to conceptualize these two pressures is as the \key{cost} and \key{informativity} of language use:
Producing and processing simple utterances incurs little cost, but more complex and diverse utterances are required to be provide higher informativity.
The idea that language results from the tension between these two pressures has a long and fruitful history.
\cite{horn-toward-1984}, RSA, Color naming

%In the domain of word order,

%TODO introduce the terms \key{cost} and \key{informativity}
%
This idea has been shown to account for phenomena such as pragmatic inference \cite{horn-toward-1984, frank2012predicting} and color naming~\cite{zaslavsky2018efficient}.

%TODO somewhere start calling this Efficiency
%TODO explain why surprisal. some production/speaker work. (plan reuse)

@@

In this work, we show that the word order of languages has evolved to optimize efficiency, and that efficiency optimization accounts for the Greenbergian word order correlations.

%In this work, we show computationally, using corpus data from 51 languages, that efficiency optimization accounts for the Greenbergian word order correlations.
That is, we show that the word orders in natural languages have evolved to optimize efficiency, and that this optimization accounts for the prevalence of the word order correlations. %word orders in natural language

@@

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|llllllll}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &         \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object     & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
	adposition    &    NP       \\
	\emph{to}            & \emph{a friend} \\
copula    &    NP        &     \\
	\emph{is}        & \emph{a friend} \\
auxiliary    &    VP       &   \\
	\emph{has}          & \emph{written} \\
noun    &    genitive      &  \\
	\emph{friend} &  \emph{of John} \\
noun    &    relative clause       &  \\
	\emph{books} & \emph{that you read} \\
complementizer    &    S        & \\
	\emph{that} & \emph{Mary} \\
verb    &    PP     &    \\
	\emph{went} & \emph{to school} \\
want    &    VP      &    \\
	\emph{wants}   &  \emph{to leave} \\
verb    &    subject    &  \\
	\emph{(there) entered} & \emph{a person} \\
verb    &    manner adverb     & \\
	\emph{ran} & \emph{quickly} \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
For each correlation, we provide an example.
Given the statistical nature of the correlations, not every language satisfies every one of them.
Not all correlations are satisfied by every natu
}\label{table:corr-dryer}
\end{table*}

@@

We model the process where a speaker communicates a message to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In Shannon's model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of entities, events, and a set of syntactic and semantic relations that hold between them.
Following a long tradition in formal and computational liguistics, we formalize these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
In Figure~\ref{fig:comm}, the message consists of a dependency structure with the words `saw', `Mary', `John', where Mary is the subject of the event denoted by `saw', and John is the object.

@@

In Information Theory, the usefulness of a communication channel depends on (1) how costly encoding and transmission is, and (2) how precisely messages can be recovered from codes.
In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance (TODO can we find an example for this?)
This reflects the \key{parseability} of the language.

The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances $u$ provide about their underlying dependency structures $t$:
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
This quantity formalizes Zipf's principle of listener effort and his Force of Diversification:
It can be maximized when trees can be decoded fully unambiguously from utterances.
Maximizing the information provided by utterances about meanings is a standard formalization of Zipf's principle of listener economy~\cite{frank2012predicting,regier2015word}.

%The counteracting Force of Unification is formalized by the cost of encoding the dependency structure and producing the utterance.
%TODO motivate surprisal here.

@@

%Do languages optimize their word order for efficiency?
%Does this optimization explain the observed language universals?

@@

% based on related work in information theory \cite{cover2006elements,tishby1999information}, bounded rationality \cite{genewein2015bounded}, and computational neuroscience \cite{friston2009predictive}; (ii) 
%
%We formalize the notion of communicative efficiency using two terms: one having to do with minimizing the difficulty of linguistic communication, and one having to do with maximizing the information transferred in such communication. We quantify the difficulty of linguistic communication in terms of the word-by-word \key{predictability} of strings, which has been found to be a highly accurate and general predictor of human online comprehension difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect},
% in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}. We quantify the information transferred in language as \key{parseability}: the quantity of information linguistic utterances contain about latent hierarchical tree structures. We also compare the results of optimizing for communicative efficiency to the results of optimizing for dependency length minimization, a previously proposed approximate metric of the amount of memory involved in language production and comprehension \cite{hudson1995measuring,gibson1998linguistic,gibson2000dependency,liu2017dependency,temperley2018minimizing}.

We formalize the cost of an utterance as its word-by-word surprisal, the degree to which it is unpredictable from the general statistics of the language.
Surprisal has been found to be is a highly accurate and general predictor human online processing difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}, and can be justified  in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}.
In expectation over all utterances $u$ in a language, the negative surprisal describes the \key{predictability}, or negative entropy, of the utteranes: %this quantity is the entropy: % of the sentences:
%The \key{predictability} of the language is the negative entropy of the sentences:
\begin{equation}
	R_{Pred} := - H[\mathcal{U}] = \sum_{u} p(u) \log p(u)
\end{equation}
where the sum runs over all possible sentences $l$ that belong to the language.
In keeping with Zipf's Force of Unification, this quantity describes how homogeneous the language is, i.e., it is larger if the distribution over sentences is concentrated on a smaller number of frequent sentences. %how many distinct sentences there are.

%It is larger

%
%The opposing pressure is \key{parsability}, which is the mutual information beteen sentences and dependency structures:
%\begin{equation}
%	R_{Pars} := I[L,T] = \sum_{t,l} p(t,l) \log \frac{p(t|l)}{p(t)}
%\end{equation}
%where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
%This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
%It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.

The \key{efficiency} of a language is a weighted combination
\begin{equation}\label{eq:efficiency}
	R_{Eff} := R_{Pars} + \lambda R_{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification).

@@

%The two terms of Eq.~\ref{eq:efficiency} implement two opposing pressures. Parseability (the mutual information term) pushes to make strings more heterogeneous, so that they can indicate different tree structures unambiguously. Predictability (the entropy term) pushes to make strings more homogeneous, to reduce the complexity of using the language.
%The idea that natural language arises from a tension between these two forces goes back over a century \cite{gabelentz1901sprachwissenschaft}.
Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking dependency structures as the underlying meanings to be conveyed.
Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship).

%\subsection{Optimizing Grammars for Efficiency}

Our goals are to show that
(1) the grammars of natural languages evolve towards optimizing efficiency of communication, and
(2) this process of optimization accounts for Greenberg's order correlation universals.

@@

We first demonstrate that real languages are relatively efficient compared to random baselines. For each language, we generated ten baseline word order grammars for the language.
These grammars have systematic word order regularities similar to natural language, but do not exhibit any correlations among the orderings of different syntactic relations (compare Figure~\ref{fig:grammars}).
%by choosing all word order grammar parameters randomly at uniform from $[0,1]$. Figure~\ref{fig:pareto-plane} shows the predictability and parseability for each real language relative to its baseline grammars. In order to control for limitations due to our word order grammar formalism, we represent real languages in the figure by maximum likelihood fits of word order grammars to the real language data.
%; in general these maximum-likelihood word order grammars end up with lower predictability and parseability than the real languages. % TODO verify this is true? TODO awkward sentence
For the calculation of predictability and parseability, we make all (baseline and real) word order grammars deterministic by always choosing the highest-probability linearization of each tree; by making the grammars deterministic in this way we eliminate an anticonservative bias toward low predictability in the baseline languages, which are highly nondeterministic. 
The majority of real languages in Figure~\ref{fig:pareto-plane} are below and to the left of their baseline equivalents, demonstrating that they are relatively high in predictability and/or parseability.

@@

\subsection{Efficiency Explains DLM}
We demonstrate the relationship between dependency length minimization and the maximization of efficiency. Figure~\ref{fig:deplength} shows average dependency length per sentence length for four typologically distinct languages, showing real languages, random baselines, and languages optimized for dependency length, parseability, predictability, and efficiency. We see that optimizing for efficiency lowers dependency length relative to random baselines, in keeping with the suggestion that dependency length minimization is a by-product of efficiency maximization \cite{futrell2017memory}. In 80\% of the languages, optimizing explicitly for dependency length produces dependencies that overshoot the dependency length of the real language; in 3/4 of the lanuages shown, the real language is best matched by efficiency optimization.

@@

We now show that efficiency optimization predicts Greenberg's word order correlations. % in optimized grammars. 
We base our evaluation on the authoritative compilation by Dryer \cite{dryer1992greenbergian}, which draws on 625 languages. %, which we take as the basis of our evaluation.
In \cite{dryer1992greenbergian}, all word order correlations are relative to the position of the direct object wrt the main verb of a sentence.
Most of them can be straightforwardly implemented in the UD formalism, allowing us to check which correlations a word order grammar satisfies.
%This dependency corresponds to dependencies of the form $\xrightarrow{obj}$ in UD.
%Similarly, most of the other dependencies in Dryer's formulation can be formalized:
%for instance, the dependency between nouns and relative clauses can be formalized as $\xrightarrow{acl}$.
%Testing whether an ordering model satisfies the correlation between relative clauses and objects reduces to checking whether these two relation types have direction parameters $a_\tau$ that are either both $>0.5$ (both kinds of dependents precede their heads) or both $< 0.5$ (they both follow their heads).

Dryer \cite{dryer1992greenbergian} presents three correlations that are only relevant to part of the world's languages and are not annotated reliably in UD. %: the dependencies between question particles and verbs, those between nouns and plural words, and those between nouns and articles.
Further collapsing correlations that cannot be formalized individually in UD, we otained 10 formalized correlations from Dryer's 15 correlations.
%Two pairs of Dryer's correlations, namely those for the dependencies between complementizers and adverbial subordinators and their complement clauses,and those for the dependencies between verbs and adpositional phrases, and adjectives and their standard of comparison, had to be collapsed into two correlations in UD. From Dryer's 15 correlations, we obtained 10 formalized correlations, roughly covering nine of Greenberg's original universals. %TODO update to 10? What happened to the 11th? Was that amod? [The 11th was adjective+standard of comparison, which is merged into obl] Maybe we could include amod as a "true rejection" of efficiency, but a false alarm for DepL

@@

See the SI for details on the neural language models and parsers used, and on the optimization procedures.}

@@

See the SI for details on the neural language models and parsers used, and on the optimization procedures.}

### 2019-02-09 second-draft/pnas-word-order-si.tex
revising paper

>>>>

\begin{figure}
    \centering
%     \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-efficiency.pdf}
 %    \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-langmod.pdf}
  %   \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-parseability.pdf}
   %  \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-depl.pdf}
     \includegraphics[scale=.45]{../results/correlations/figures/posterior-satisfied-universals-together.pdf}

	\caption{Posterior of the number of correlations predicted in the direction found in the UD treebanks, computed from a mixed-effects logistic regression jointly modeling all ten dependencies. Efficiency predicts all ten correlations with high posterior probability.}
    \label{fig:posterior-satisfied-count}
\end{figure}

We further evaluate the word order predictions of efficiency, showing that efficiency is most successful in predicting correlations in the direction found in the UD corpora.
We constructed a single logistic model predicting, for each of the ten dependencies, whether it is correlated or anti-correlated with the \emph{obj} dependency in languages optimized for efficiency, with random effects for language and language family, correlated across the ten dependencies.
We conducted the same analysis for predictability, parseability, and dependency length.
We used this model to estimate the posterior distribution of the number of correlations that an objective function predicts to be in the same direction as found in the UD treebanks.
The resulting distributions are shown in Figure~\ref{fig:posterior-satisfied-count}.
The estimated posterior probability that efficiency predicts less than all ten dependencies to correlate in the same direction as in the UD treebanks is $0.0242$. The probability that it predicts less than nine of the correlations is $3\cdot 10^{-4}$.
For dependency length, the posterior puts much of the probability mass on predicting only nine of the correlations; predictability and parseability predict significantly less correlations.
